---
title: "Spatial distribution of dog bite incidents and dog licenses"
output:
  html_document:
    code_folding: hide
    always_allow_html: true
    toc: true
    toc_float: true
---

# Overview of Dog Bite Count and License Count in NYC

The following map displays the spatial distribution of dog licenses and dog bite incidents across various zip codes in New York City from 2015 to 2022. The color fill of the map indicates the number of dog licenses in each zip code, while the dots represent the total number of dog bite incidents in each area.

The preliminary findings of the study suggest that the geographic distribution of dog bite incidents exhibits spatial autocorrelation, with certain zip codes, particularly in Midtown Manhattan, identified as high-frequency areas for such incidents.

Additionally, the data indicates that areas with a higher number of dog licenses, such as zip codes 10025, 10024, and 10314, typically have higher numbers of dog bite incidents.   Therefore, there may be a positive correlation between the number of dog licenses and dog bite incidents.

<iframe src="maps/Dog_Bite_Count_Switchable.html" style="width: 100%; height: 700px; border: none;"></iframe>

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(rvest)
library(broom)
library(dplyr)
library(purrr)
library(readxl)
library(plotly)

```

```{r data, include=FALSE, message=FALSE}
url = "https://p8105.com/data/zip_codes.html"
ny_zip_codes = read_html(url) |>
  html_table() |> 
  data.frame() |> 
  janitor::clean_names() |> 
  mutate(
    borough = factor(
      county,
      levels=c("Bronx","Kings","New York", "Queens","Richmond"),
      labels=c("Bronx","Brooklyn","Manhattan","Queens",
 "Staten Island"))) 

dog_bites_df = read_csv("data/Dog_Bites_Data.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("^x", "", .))

# glimpse(dog_bites_df)
# summary(dog_bites_df)
unique_breeds <- dog_bites_df %>%
  pull(breed) %>%
  unique() %>%
  paste(collapse = ", ")

dog_bites_clean = dog_bites_df |> 
  mutate(date_of_bite = as.Date(date_of_bite, format = "%B %d %Y")) |> 
  mutate(year = format(date_of_bite, "%Y"),
         month = format(date_of_bite, "%m"),
         day = format(date_of_bite, "%d")) |> 
  mutate(year = as.factor(as.numeric(year)),
         month = factor(as.numeric(month), levels = 1:12),
         day = as.factor(as.numeric(day))
         ) |> 
  mutate(breed = str_to_lower(breed))

dog_licensing_df = read_csv("data/NYC_Dog_Licensing_Dataset.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  rename_with(~ gsub("^x", "", .))
```

# Relationship between dog licenses and dog bites

```{r message=FALSE, warning=FALSE}
valid_zipcodes = pull(ny_zip_codes, zip_code)
bites_by_zip = dog_bites_clean |>
  group_by(zip_code) |>
  summarise(Bite_Count = n())

licenses_by_zip = dog_licensing_df |>
  group_by(zip_code) |>
  summarise(License_Count = n())
zip_data = merge(bites_by_zip, licenses_by_zip, by = "zip_code", all = TRUE) |> 
  mutate(
    zip_code = as.factor(zip_code)
  ) |> 
  filter(zip_code %in% valid_zipcodes) |> 
  filter(!is.na(License_Count), !is.na(Bite_Count))

scatter_plot_interactive = ggplotly(
  ggplot(data = zip_data, aes(x = License_Count, y = Bite_Count)) +
    geom_point(color = "blue", alpha = 0.6) +
    geom_smooth(method = "lm", color = "red") +
    labs(title = "Relationship", x = "licenses", y = "Bite") +
    theme_minimal()
)

scatter_plot_interactive
```



